---
title: "Permutation Test"
author: Elene Barbakadze
date: October 28, 2025
format: 
  html:
    code-fold: true
    code-summary: "Show code"
execute: 
  warning: false
  message: false
---

The dataset, Students Data Analysis, contains information on students’ graduation acceptances (did not get into the graduate school, got into a local graduate school, got into a international graduate school) and their assigned class (class), a categorical variable with two groups. Students were initially divided into two classes, but instructors suspected that students in one class might be performing significantly differently from those in the other.

To test this hypothesis, I will use a permutation test to compare acceptance rates between the two classes. The procedure involved repeatedly shuffling the class labels while keeping the graduation scores fixed, then calculating the mean difference in graduation rates for each permutation. This generated a null distribution representing the differences we would expect if class assignment had no real effect on performance.

IMPORTANT NOTE: The dataset used in this analysis is half-synthetic, meaning that although it was originally based on student records, the variables have been anonymized, censored, and partially generated to protect privacy. As a result, the class categories do not represent real student groups, and the patterns observed in the data may not reflect any real-world population. The analysis here should therefore be interpreted purely as a practice exercise for understanding and applying permutation tests.

### Hypotheses

**Null hypothesis (H₀):** There is no difference in graduate school acceptance rates between the two classes. Any observed difference is due to random chance.

**Alternative hypothesis (Hₐ):** There is a difference in graduate school acceptance rates between the two classes.

Loading the dataset and packages:

```{r}
library(RTextTools) 
library(tidyverse)
library(stringr)
library(readr)
library(purrr)
library(ggplot2)
library(gt)
students <- read.csv("Students Data.csv")

```

### Step 1: Explore and Summarize the Dataset

Because the dataset includes two types of graduate school acceptances (international and local) I combined them into a single category and calculated the average graduation acceptance rate.

```{r}
# Create binary grad variable and keep class and grad columns
students_clean <- students |> 
  mutate(grad = as.integer(y>0)) |> 
  select(class, grad) 

# Calculate acceptance proportion by class and display table
students_clean |> 
  group_by(class) |> 
  summarize(ave_grad = mean(grad)) |> 
  gt() |> 
  tab_header(
    title = "Average Grad School Acceptances by Class"
  ) 

```

```{r}
# Compute difference in acceptance proportions between the two classes
students_clean |> 
  group_by(class) |> 
  # proportion accepted per class
  summarize(ave_grad = mean(grad)) |> 
  # difference: Class B - Class A
  summarize(ave_diff =diff(ave_grad)) |> 
  gt() |>
  tab_header(
    title = "Differences in Grad School Acceptances Between Classes"
  )
```

From the analysis, we can see that Class A had a higher graduate school acceptance rate than Class B.

Before conducting the permutation test, we can visualize the proportion of students accepted to graduate school in each class. This helps contextualize the observed difference in acceptance rates.

```{r}
#| fig-cap: Proportion of students accepted to graduate school in each class.
#| fig-alt: Bar chart comparing the proportion of students accepted to graduate school between Class A and Class B. Class A’s bar is taller, indicating approximately 51 percent acceptance, while Class B shows about 38 percent. This visual summarizes the observed difference before conducting the permutation test.

# Show proportion (mean) of grad acceptance by class
students_clean |> 
  group_by(class) |> 
  summarize(prop_accept = mean(grad)) |> 
  ggplot(aes(x = class, y = prop_accept, fill = class)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Students Accepted to Graduate School by Class",
    x = "Class",
    y = "Proportion Accepted"
  ) +
  theme_light()
```

### Step 2: Permutation Test Function

This function tests whether the difference in the **proportion of students accepted to graduate school** between the two classes could have occurred by chance alone. Because grad is coded as 0 or 1, the average within each class represents the **proportion** accepted. The test works by shuffling the grad values across students, recalculating the difference in acceptance proportions for each permutation, and comparing those values to the observed difference. This produces both the observed statistic (ave_diff) and a distribution of permuted statistics (perm_diff). I set a random seed to ensure reproducibility and used map() to repeat the permutation process 1,000 times, storing all results in a data frame called perm_stats.

```{r}
#| code-fold: show
#| echo: true

#a function that computes the observed and permuted differences
grad_diff <- function(rep, students_clean) {
  students_clean |> 
    # shuffle grad values
    mutate(grad_perm = sample(grad, replace = FALSE)) |> 
    group_by(class) |> 
    summarize(
       # observed proportion
      obs_ave = mean(grad),
      # permuted proportion
      perm_ave = mean(grad_perm)) |> 
    summarize(
      # observed difference
      ave_diff =diff(obs_ave), 
      # permuted difference
      perm_diff = diff(perm_ave), 
      # iterate label
      rep = rep)
}

# Run the permutation test 1000 times
set.seed(47)
number <- 1000
perm_stats<- map(c(1:number), grad_diff, students_clean) |> 
  list_rbind()

```

### Step 3: Visualize the Null Distribution

The histogram below displays this null distribution of permuted mean differences (perm_diff), with the red line showing the observed mean difference from the original data (ave_diff).

```{r}
#| fig-cap: Permutation distribution of the difference in graduate school acceptance proportions between the two classes.
#| fig-alt: Histogram showing the distribution of permuted differences in graduate school acceptance proportions generated from 1,000 random shuffles of the outcome variable. The values cluster around zero, as expected under the null hypothesis. A red line indicates the observed difference from the actual data, which lies within the central region of the permutation distribution, suggesting that such a difference could plausibly occur by chance.

perm_stats |> 
  # histogram of permuted differences
  ggplot(aes(x = perm_diff))+
  geom_histogram(bins  = 40 ,fill = "skyblue", color = "white", alpha = 0.8) +
  # Add title and axis labels
  labs(
    title = "Permutation Distribution of Mean Differences",
    subtitle = "Red line = observed difference",
    x = "Permutation Mean Difference",
    y = "Count"
  ) +
  # mark observed difference
  geom_vline(aes(xintercept = ave_diff), color = "red", linewidth = 1.2)+
  theme_classic()+
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

From the plot, we can see that the red line, which represents the observed difference, is slightly to one side of the null distribution. However, it does not seem far enough from the center to suggest a meaningful or statistically significant difference at first glance.

### Step 4: Calculate the p-value

Finally, I calculated the p-value

```{r}
# Compute the two-sided p-value:
perm_stats |> 
  summarize(p_val_ave = mean(abs(perm_diff) >= abs(ave_diff))) |>
  gt() |> 
  tab_header(
    title = "Left-tailed p-value"
  )
```

Because this is a two-sided test, the p-value was computed as the proportion of permuted differences whose absolute value was greater than or equal to the absolute observed difference.

### Step 5: Results

The permutation test produced a p-value greater than 0.05, meaning the observed difference in graduate school acceptance proportions between the two classes is not statistically significant. In other words, the difference we see could easily have arisen by random chance, so we fail to reject the null hypothesis (H₀). Based on this analysis, there isn’t enough evidence to conclude that class assignment meaningfully affects graduate school acceptance within this dataset.

**Citation:**\

Zhou, Erqi. *Students Data Analysis.* Synthetic dataset created from anonymized student records.

Zhou, Erqi. *Students Data Analysis.* Kaggle, 2024. <https://www.kaggle.com/datasets/erqizhou/students-data-analysis>.
