---
title: "Data Wrangling with SQL"
author: Elene Barbakadze
date: December 1, 2025
format: 
  html:
    code-fold: true
    code-summary: "Show code"
execute: 
  warning: false
  message: false
---

### Introduction

In this analysis, I examine traffic stop data from three cities to understand how race and gender shape who gets stopped. All data wrangling is done in SQL, including extracting racial distributions, proportions of stops within each city, and demographic patterns across groups. I also incorporate external population data to compare stop patterns to the cities’ underlying demographic makeup. The final plots and tables highlight clear differences across cities and provide a basis for interpreting how stop patterns vary by group.

### Data Wrangling

Establishing the connection:

```{r}

con_traffic <- DBI::dbConnect(

  RMariaDB::MariaDB(),

  dbname = "traffic",

  host = Sys.getenv("TRAFFIC_HOST"),

  user = Sys.getenv("TRAFFIC_USER"),

  password = Sys.getenv("TRAFFIC_PWD")

)
```

Loading libraries:

```{r}
library(DBI)
library(dplyr)
library(ggplot2)
```

Before diving into the SQL analysis, it’s useful to look at the overall size of the dataset across the three cities. The bar chart below shows the total number of traffic stops recorded in Los Angeles, Austin, and Nashville. Seeing the relative scale of each dataset helps provide context for the later comparisons by race and gender, since the number of stops varies considerably from city to city.

```{sql, connection = con_traffic, output.var = "stops_by_city"}

# Total stops in Los Angeles
SELECT 'Los Angeles' AS city, COUNT(*) AS num_stops
FROM ca_los_angeles_2020_04_01

UNION ALL

# Total stops in Austin
SELECT 'Austin' AS city, COUNT(*) AS num_stops
FROM tx_austin_2020_04_01

UNION ALL

# Total stops in Nashville
SELECT 'Nashville' AS city, COUNT(*) AS num_stops
FROM tn_nashville_2020_04_01;

```

```{r}
#| fig-cap: A bar chart showing total number of traffic stops recorded in Los Angeles, Austin, and Nashville.

#| fig-alt: Bar chart showing the total number of traffic stops in three cities: Los Angeles has the highest count, Nashville has the second highest, and Austin has the smallest dataset.

# Create a bar chart showing total stops per city
stops_by_city |>
  ggplot(aes(x = city, y = num_stops, fill = city)) +
  # Titles and axis labels, remove legend
  geom_col(show.legend = FALSE) +
  labs(
    title = "Total Number of Traffic Stops by City",
    x = "City",
    y = "Number of Stops"
  ) +
  theme_bw()
```

Extracting the proportions of stops in Los Angeles, Austin, and Nashville by race:

```{sql, connection=con_traffic, output.var="city_race_prop"}

# Compute the number of stops for each race in each city
SELECT
  race_count.subject_race,
  race_count.city,
  race_count.n,
  race_count.n * 1.0 / total.total AS prop
FROM (
  -- Count stops by race in Los Angeles
    SELECT 
        subject_race,
        COUNT(*) AS n,
        'Los Angeles' AS city
    FROM ca_los_angeles_2020_04_01
    GROUP BY subject_race

    UNION ALL

    -- Count stops by race in Austin
    SELECT 
        subject_race,
        COUNT(*) AS n,
        'Austin' AS city
    FROM tx_austin_2020_04_01
    GROUP BY subject_race

    UNION ALL

    -- Count stops by race in Nashville
    SELECT 
        subject_race,
        COUNT(*) AS n,
        'Nashville' AS city
    FROM tn_nashville_2020_04_01
    GROUP BY subject_race

) AS race_count

-- Compute total stops per city (needed to calculate proportions)
JOIN (
    SELECT 'Los Angeles' AS city, COUNT(*) AS total
    FROM ca_los_angeles_2020_04_01

    UNION ALL

    SELECT 'Austin', COUNT(*)
    FROM tx_austin_2020_04_01

    UNION ALL

    SELECT 'Nashville', COUNT(*)
    FROM tn_nashville_2020_04_01
) AS total
ON race_count.city = total.city
-- Sort
ORDER BY race_count.city, race_count.subject_race;
```

I used the join because the race counts and the total counts require two different levels of aggregation. The race_count subquery groups by race, while the total subquery counts all stops in each city. Joining them lets me compute n / total cleanly. Since MySQL on Posit doesn’t allow window functions in this setup, the join is the simplest valid approach.

Extracting the number of stops in Austin based on gender:

```{sql, connection=con_traffic, output.var="gender_austin"}

-- the number of stops in austin based on gender

-- Count stops by race for female subjects
(SELECT 
    subject_race, 
    count(*) as n, 
    "female" as sex
  FROM tx_austin_2020_04_01
  WHERE subject_sex = "female"
  GROUP BY subject_race)
  
  UNION ALL

-- Count stops by race for male subjects
(SELECT 
    subject_race, 
    count(*) as n,
    "male" as sex
  FROM tx_austin_2020_04_01
  WHERE subject_sex = "male"
  GROUP BY subject_race)
  
-- Sort
ORDER BY subject_race, sex
```

Extracting the average number of people who were stopped in Austin by gender:

```{sql, connection=con_traffic, output.var="gender_age_austin"}

-- number of stops in austin based on gender and age

-- Average age by race for female subjects
(SELECT 
    subject_race, 
    AVG(subject_age) as age, 
    "female" as sex
  FROM tx_austin_2020_04_01
  WHERE subject_sex = "female"
  GROUP BY subject_race)
  
  UNION ALL

-- Average age by race for male subjects
(SELECT 
    subject_race, 
    AVG(subject_age) as age,
    "male" as sex
  FROM tx_austin_2020_04_01
  WHERE subject_sex = "male"
  GROUP BY subject_race)
  -- Sort
ORDER BY subject_race, sex
```

To compare stop rates with the underlying demographics of each city, I used racial population estimates from the 2020 U.S. Census, accessed through the CensusDots database. The Census provides city-level counts for major racial categories, which I matched to Los Angeles, Austin, and Nashville. These population values reflect the residential makeup of each city in 2020, allowing me to compare the racial distribution of stopped drivers to the racial distribution of the city’s population in the same period.

Dataframe with the population distribution in Los Angeles, Austin, and Nashville:

```{r}
population_data <- data.frame(
  city = c(
    "Los Angeles", "Los Angeles", "Los Angeles", "Los Angeles", "Los Angeles",
    "Austin",  "Austin",   "Austin",   "Austin",   "Austin", 
    "Nashville",  "Nashville",  "Nashville",  "Nashville", "Nashville"
  ),
  subject_race = c(
    "asian/pacific islander",
    "black",
    "hispanic",
    "other",
    "white",

    "asian/pacific islander",
    "black",
    "hispanic",
    "other",
    "white",

    "asian/pacific islander",
    "black",
    "hispanic",
    "other",
    "white"
  ),
  n = c(
    14.7, 7.6, 	48, 41, 25.6,    #Los Angeles
    8.9, 6.9, 32.5, 4.7, 47.1,    # Austin
    3.9, 24.3, 14, 4.5, 53.3    # Nashville
  )
)

```

### Data Visualisation

```{r}
#| fig-cap: Bar graph of proportion of traffic stops by race across Austin, Los Angeles, and Nashville.
#| fig-alt: Bar chart faceted by city showing the proportion of traffic stops by race. In Austin and Nashville, white drivers make up the largest share of stops. Los Angeles shows a higher proportion of Hispanic drivers. Black drivers represent a substantial share in all three cities, especially Nashville. Asian/Pacific Islander and Other categories appear in much smaller proportions.

# Plot the proportion of traffic stops by race for each city.
city_race_prop |> 
  filter(!is.na(subject_race), subject_race != "unknown") |> 
  ggplot(aes(x = subject_race, y = prop, fill = subject_race)) +
  geom_col() +
  # Using facet_wrap lets us compare racial distributions across cities.
  facet_wrap(~ city) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  ) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Distribution of Stopped Individuals by Race",
    y = "Proportion of stops",
    fill = "Race of Subject"
  )
```

This graph shows the distribution of individuals stopped by the police in 3 cities – Los Angeles, Austin, and Nashville. Looking at the graph, there seems to be significant race differences – for example, in Austin and Los Angeles, it is predominately white and hispanic individuals who are stopped. In Nashville, the largest groups are white and Black individuals. However, we should be careful when it comes to arriving to conclusions from this graph. To understand the full picture, we need to compare these stop patterns to the general racial makeup of each city.

```{r}
#| fig-cap: Population distribution by race across Austin, Los Angeles, and Nashville.
#| fig-alt: A set of three bar charts comparing racial population distributions in Austin, Los Angeles, and Nashville. Each panel shows the proportion of residents identified as Asian or Pacific Islander, Black, Hispanic, Other, and White. Austin’s largest group is White, followed by Hispanic. Los Angeles shows a majority Hispanic population with a substantial “Other” category. Nashville has a majority White population with a notable Black population. The colors for each race category match those used in the traffic stop plot for easy comparison.

 # population counts by race for each city
population_data |> 
  ggplot(aes(x = subject_race, y = n, fill = subject_race)) +
  geom_col() +
  facet_wrap(~ city) +
  theme_bw() +
  # Remove x-axis labels
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  ) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Population Distriburion by Race",
    y = "Percentage",
    fill = "Race of Subject"
  )
```

Once we compare the stop data to the general population, the picture gets clearer. In Austin, the racial breakdown of stopped individuals closely mirrors the city’s demographics. Los Angeles shows a different pattern, where Asian and Pacific Islander individuals appear overrepresented in stops, and people classified as “other” are underrepresented. In Nashville, the proportion of stopped Black individuals is noticeably higher than their share of the population, suggesting a potential racial bias in policing.

```{r}
#| fig-cap: Distribution of traffic stops in Austin by race and sex.

#| fig-alt: A set of two bar charts comparing traffic stops in Austin by race, split into female and male subjects. Each chart shows counts for five racial groups: Asian or Pacific Islander, Black, Hispanic, Other, and White. White individuals have the highest number of stops in both groups, followed by Hispanic and Black individuals, with Asian/Pacific Islander and Other making up much smaller proportions.

# Plot number of stops in Austin by race, separated by sex
gender_austin |> 
  filter(!is.na(subject_race),subject_race != "unknown") |> 
  ggplot(aes(x = subject_race, y = n, fill = subject_race)) +
  geom_col() +
  facet_wrap(~ sex) +
  theme_bw() +
  # Remove x-axis labels
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  ) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Distribution of Stopped Individuals in Austin by Race",
    y = "Number of stops",
    fill = "Race of Subject"
  )
```

For the next part of the analysis, I looked at how the racial breakdown of stops changes when we split the data by gender. I focused on Austin for this comparison. Overall, the racial patterns look similar across men and women, but the total number of stops is consistently higher for men.

```{r}
#| fig-cap: Average age of stopped individuals in Austin, separated by race and sex.

#| fig-alt: A faceted bar chart showing the average age of stopped individuals in Austin. The left panel shows females and the right panel shows males. For each sex, bars represent racial groups: Asian/Pacific Islander, Black, Hispanic, Other, and White. Ages mostly fall between 34 and 40, with male subjects generally slightly older across all groups, and white males having the highest average age.

# Plot average age by race, separated by sex
gender_age_austin |> 
  filter(!is.na(subject_race),subject_race != "unknown") |>
  ggplot(aes(x = subject_race, y = age, fill = subject_race)) +
  geom_col() +
  facet_wrap(~ sex) +
  theme_bw() +
  # Remove x-axis labels 
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  ) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Average Age of Stopped Individual in Austin",
    y = "Average Age of a subject",
    fill = "Race of Subject"
  )+
  # Limit y-axis to 30–40
  coord_cartesian(ylim = c(30, 40))
```

I also examined the average age of stopped individuals in Austin by both race and gender. Among women, Black individuals had the highest average age, followed by white individuals. For men, this pattern flips, with white individuals having the highest average age instead. Overall, the average age is slightly lower for women, ranging from 33.6 (unknown) to 36.5 (Black), compared to men, whose averages range from 35.7 (Asian/Pacific Islander) to 39.8 (white).

### Conclusion

Terminating SQL connection:

```{r}
dbDisconnect(con_traffic)
```

Overall, this project combined SQL wrangling, external population data, and data visualization to explore how demographic factors show up in traffic stops. The goal was to move from simple counts to a more contextual understanding by comparing stop patterns to who actually lives in each city. Focusing on Austin also showed how gender and age add another layer to these differences.

### References

CensusDots. “Austin, TX Demographics.” <https://www.censusdots.com/race/austin-tx-demographics>.

CensusDots. “Los Angeles County, CA Demographics.” <https://www.censusdots.com/race/los-angeles-county-ca-demographics>.

CensusDots. “Nashville–Davidson Metropolitan Government Balance, TN Demographics.” <https://www.censusdots.com/race/nashville-davidson-metropolitan-government-balance-tn-demographics>.

Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, et al. 2020. “A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States.” *Nature Human Behaviour*, 1–10.
